---
title: |
  Project Part 1
  
  **Maternal Mortality Rates in the United States**
author: Shanyar Ghaderi
fontsize: 12pt
geometry: margin=1in
urlcolor: black
output: pdf_document
header-includes:
- \usepackage{setspace}
- \onehalfspacing
---

\noindent GitHub Repo: \url{https://github.com/sghaderi24/STAT3080_Maternal_Mortality}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, highlight=FALSE)
```

## Introduction

Maternal mortality is a critical public health issue in the United States, where pregnancy-related deaths occur at higher rates than in other developed countries. Understanding trends in maternal mortality is essential for identifying potential disparities and improving healthcare outcomes.

This project analyzes maternal mortality rates using national data from the Centers for Disease Control and Prevention (CDC). The goal is to explore patterns in mortality rates over time.

The research questions are:

1. What is the average maternal mortality rate in the United States?
2. How has maternal mortality changed over time?
3. What patterns or trends are observable in maternal mortality rates?

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# -------------------------
# Load data
# -------------------------
library(tidyverse)

data <- read.csv("maternal_mortality_us_clean.csv")

names(data)
data <- data %>%
  rename(
    Year = year,
    MortalityRate = maternal_mortality_rate_per_100k
  )

data <- data %>%
  filter(!is.na(MortalityRate))
```

## Data Summary

The dataset used in this analysis was obtained from the National Center for Health Statistics (NCHS), a division of the CDC. The dataset reports maternal mortality rates per 100,000 live births in the United States.

(a) Population vs Sample

The data represent a population dataset, as they include all reported maternal deaths in the United States rather than a subset or sample.

(b) Data Collection

The data were collected through the National Vital Statistics System (NVSS), which compiles information from death certificates across all U.S. states. Maternal deaths are identified using standardized definitions related to pregnancy-related causes.

(c) Data Modifications

Several preprocessing steps were applied to prepare the dataset for analysis:

- Removal of missing values
- Standardization of column names
- Selection of relevant variables (Year and Mortality Rate)

These changes ensure consistency and allow for accurate analysis.

(d) Potential Issues

There are several limitations associated with this dataset:

- Maternal deaths may be underreported or misclassified
- Changes in reporting practices over time may affect comparability
- External factors such as socioeconomic status and healthcare access are not included

These limitations may impact the interpretation of trends.

(e) Appropriateness of Data

The dataset is appropriate for this analysis because it directly measures maternal mortality rates, which align with the research questions. Additionally, the use of national-level data improves the reliability and generalizability of the findings.


## Exploratory Analysis
```{r, echo=FALSE}
# -------------------------
# Numerical summaries
# -------------------------
cat("Mean maternal mortality rate:", mean(data$MortalityRate), "\n")
cat("Standard deviation:", sd(data$MortalityRate))

# -------------------------
# Graph 1: Line plot
# -------------------------
ggplot(data, aes(x = Year, y = MortalityRate)) +
  geom_line() +
  geom_point() +
  labs(title = "Maternal Mortality Rate Over Time",
       x = "Year",
       y = "Deaths per 100,000 Live Births")

# -------------------------
# Graph 2: Histogram
# -------------------------
ggplot(data, aes(x = MortalityRate)) +
  geom_histogram(bins = 10) +
  labs(title = "Distribution of Maternal Mortality Rates",
       x = "Mortality Rate",
       y = "Frequency") 
```

## Conclusions

The exploratory analysis provides insight into trends and variability in maternal mortality rates in the United States.

The mean maternal mortality rate represents the average number of maternal deaths per 100,000 live births across the observed time period, while the standard deviation reflects the degree of variation in these rates from year to year. These numerical summaries indicate that maternal mortality is not constant and varies across time.

The line plot reveals a general upward trend in maternal mortality rates in recent years, suggesting that maternal health outcomes may be worsening. The histogram shows the distribution of mortality rates, indicating that higher values occur more frequently in recent periods compared to earlier years.

Together, these findings suggest that maternal mortality has increased over time and that there is noticeable variability in the data. These patterns highlight the importance of further statistical analysis to determine whether the observed trends are statistically significant and to better understand the factors contributing to these outcomes.

\newpage
## References

Hoyert DL. Maternal mortality rates in the United States, 2024. Hyattsville, MD: National Center for Health Statistics; 2026. Available from: https://dx.doi.org/10.15620/cdc/174651

National Center for Health Statistics. Maternal mortality rates in the United States. Centers for Disease Control and Prevention. Available from: https://www.cdc.gov/nchs/data/hestat/hestat113.htm

Centers for Disease Control and Prevention. Pregnancy Mortality Surveillance System. Available from: https://www.cdc.gov/maternal-mortality/php/pregnancy-mortality-surveillance-data/index.html

Centers for Disease Control and Prevention. National Vital Statistics System: Mortality Data. Available from: https://www.cdc.gov/nchs/nvss/deaths.htm

AI assistance (ChatGPT) was used to support data analysis, coding, and report structuring in accordance with UVA guidelines.
